On Optimality of Finite Element Integration

### 2015-09-25 TOMS_Paper/assembly.tex
Fixing a few things

\section{Code Generation}
\label{sec:codegen}
The model described in Section~\ref{sec:optimal-synthesis} has been fully automated in COFFEE, the optimizer of finite element integration routines used in Firedrake. In this section, we describe the features of this code generation system.

\subsection{Automation through the COFFEE Language}
As opposed to what happens in the FEniCS Form Compiler with quadrature and tensor modes, there are no separate trunks in COFFEE handling pre-evaluation, sharing, and code motion in general. All optimizations are instead expressed as composition of parametric ``building-block'' operations. This has several advantages. Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could be expressed using the existing operators, or with small effort building on what is already available. Secondly, generality: other domains sharing properties similar to that of finite element integration (e.g., multilinear loop nests) could be optimized through the same compiler. Thirdly, robustness: the same building-block operations are exploited, and therefore stressed, by different optimization pipelines.

A non-exhaustive list of such operations includes expansion, factorization, re-association, generalized code motion. These ``rewrite operators'' can be seen as the COFFEE language. They define parametric transformations: for example, one could ask to factorize constant rather linear terms, while hoisting could be driven by loop dependency. Their implementation is based on manipulation of the abstract syntax tree representing the integration routine.

\subsubsection{Heuristic Optimization of Integration-dependent Expressions}
\label{sec:heuristic-int-dep}
As a proof-of-concept of our generality claim, we briefly discuss our optimization strategy for integration-dependent expressions. These are expressions that should logically be placed within $L_i$. They can originate, for example, from the extensive use of tensor algebra in the derivation of the weak variational form or from the use of a non-affine reference-to-physical element mapping, which Jacobian needs be re-evaluated at every quadrature point. For some complex monomials and for coarser discretizations, the operation count within $L_i$ could be comparable or, in some circumstances, even outweigh that of the multilinear loop nest. In these cases, our optimality model becomes weaker, since its underlying assumption is that the bulk of the computation is carried out in innermost loops.

Despite the fact that we are not characterizing optimality for this wider class of problems, we can still heuristically apply the same reasoning of Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis} to try to eliminate sharing, thus reducing the operation count. This is straightforward in our code generation system by composing rewrite operators. Our strategy is as follows.

>>>>

\section{Code Generation}
\label{sec:codegen}
The transformation algorithm described in Section~\ref{sec:optimal-synthesis} has been implemented in COFFEE, the optimizer of finite element integration routines used in Firedrake. In this section, we describe several aspects related to code generation.

\subsection{Automation through the COFFEE Language}
As opposed to what happens in the FEniCS Form Compiler, in which quadrature and tensor modes are completely separated code modules, COFFEE implements all optimizations by composing ``building-block'' transformations, which we refer to as ``rewrite operators''. This has several advantages. Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could be expressed using the existing operators, or with small effort building on what is already available. Secondly, generality: any sort of codes characterized by (linear, but also non-linear) loop nests with sharing can be optimized (obviously, without guarantees on optimality). COFFEE exploits domain properties, although is not specifically tied to finite element in any way. Thirdly, robustness: the same operators are exploited, and therefore stressed, by different optimization pipelines.

A non-exhaustive list of such operators includes expansion, factorization, re-association, generalized code motion. The rewrite operators can be seen as the COFFEE language. They are flexible: for example, one could ask to factorize constants instead of linear terms. Their implementation is based on manipulation of the abstract syntax tree representing the integration routine.

\subsubsection{Heuristic Optimization of Non-linear Loops}
\label{sec:heuristic-int-dep}
As a proof-of-concept of our generality claim, we briefly discuss our optimization strategy for non-linear loops. Non-linear loops (expressions) are the natural result of non-linear coefficients in the form. The non-linear expressions logically depend on $L_i$. In complex monomials and/or with coarser discretizations, the operation count can experience a significant increase due to non-linear terms.

Despite the fact that we are not characterizing optimality for this wider class of problems, we still heuristically apply the same reasoning of Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis} for (potentially) partial sharing elimination, thus reducing the operation count. This is straightforward in our code generation system by composing rewrite operators. Our strategy is as follows.
